---
title: "Predicting Mortality in Heart Failure Patients"
author: "Michael Matyevich"
date: "2026-04-06"
output:
  html_document:
    toc: yes
    toc_float: yes
    theme: flatly
    highlight: tango
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5)
library(tidyverse)
library(caret)
library(MASS)
library(pROC)
library(corrplot)
library(glmnet)
library(dplyr)

hfc_records <- read.csv("C:/Users/micha/Downloads/heart_failure_clinical_records_dataset.csv",
                        stringsAsFactors = TRUE)
```

---

## Background

Heart failure is a condition where the heart cannot pump enough blood to meet the body's needs.
It is a leading cause of hospitalisation and death worldwide, and identifying which patients are
at highest risk of dying allows clinicians to prioritise monitoring and intervention resources
more effectively.

This analysis uses clinical records from 299 heart failure patients to build and compare
classification models that predict 30-day mortality. All four models are evaluated under the
same cross-validation framework so performance comparisons are fair.

---

## Research Questions

**RQ1 — Which clinical measurements are most strongly associated with mortality in heart failure
patients?**

Understanding which variables matter most guides both clinical triage and future data collection
priorities. If a small number of cheap, routine measurements can predict mortality accurately,
that has immediate practical value.

**RQ2 — Does LASSO variable selection produce a more parsimonious model without meaningfully
sacrificing predictive performance compared to a full model?**

Simpler models are easier to explain to clinicians, less likely to overfit on small datasets,
and cheaper to implement in practice. LASSO is a principled way to find a minimal sufficient
predictor set — we compare it directly to the full model.

**RQ3 — Do linear discriminant methods (LDA, QDA) perform competitively with logistic
regression for this classification task?**

LDA and QDA make different assumptions about the data compared to logistic regression. Testing
all three helps us understand whether the distributional assumptions underlying each method
are supported by this dataset.

---

## Dataset Description

The dataset contains **299 hospitalised heart failure patients** with 13 variables recorded
during follow-up. The outcome is `DEATH_EVENT` — whether the patient died during the
follow-up period (1 = died, 0 = survived).

| Variable                  | Type        | Description                                         |
|---------------------------|-------------|-----------------------------------------------------|
| age                       | Numeric     | Age at time of hospitalisation (years)              |
| anaemia                   | Binary      | Whether the patient has anaemia (0/1)               |
| creatinine_phosphokinase  | Numeric     | Level of CPK enzyme in blood (mcg/L)                |
| diabetes                  | Binary      | Whether the patient has diabetes (0/1)              |
| ejection_fraction         | Numeric     | % of blood leaving heart per contraction            |
| high_blood_pressure       | Binary      | Whether the patient has hypertension (0/1)          |
| platelets                 | Numeric     | Platelet count (kiloplatelets/mL)                   |
| serum_creatinine          | Numeric     | Level of creatinine in blood (mg/dL)                |
| serum_sodium              | Numeric     | Level of sodium in blood (mEq/L)                    |
| sex                       | Binary      | Biological sex (0 = female, 1 = male)               |
| smoking                   | Binary      | Whether the patient smokes (0/1)                    |
| time                      | Numeric     | Follow-up period in days                            |
| DEATH_EVENT               | Binary      | **Outcome** — died during follow-up (1 = yes)       |

The class split is approximately **67% survived and 33% died** — a moderate imbalance that
means a naive model predicting "survived" for everyone would achieve 67% accuracy without
learning anything. This is why we use metrics beyond simple accuracy (sensitivity, specificity,
AUC-ROC) to evaluate performance.

---

## Exploratory Analysis

```{r hfc}
str(hfc_records)
summary(hfc_records)

hfc_records$DEATH_EVENT <- as.factor(hfc_records$DEATH_EVENT)
colSums(is.na(hfc_records))
```

There are **no missing values** in this dataset, so no imputation is required.

```{r scale}
# Scale numeric variables so LDA / QDA will have equal weighting.
# Without scaling, variables measured on larger scales (e.g. creatinine_phosphokinase
# which can reach tens of thousands) would dominate the distance calculations
# that LDA and QDA rely on, simply because of their units rather than their
# actual predictive value.
numeric_vars <- c("age", "creatinine_phosphokinase", "ejection_fraction",
                  "platelets", "serum_creatinine", "serum_sodium", "time")
hfc_records[numeric_vars] <- scale(hfc_records[numeric_vars])

table(hfc_records$DEATH_EVENT)
```

The class distribution confirms the 67/33 split noted above. This is worth keeping in mind
throughout — in a clinical context, the *cost of a false negative* (predicting "survived"
when the patient actually dies) is much higher than the cost of a false positive. We will
pay particular attention to **sensitivity** (the model's ability to correctly identify deaths)
when comparing models.

```{r corrplot}
cor_matrix <- cor(hfc_records |> dplyr::select(-DEATH_EVENT))
corrplot(cor_matrix)
```

The correlation matrix shows mostly weak relationships between predictors, which is good —
it means we are unlikely to have severe multicollinearity issues where two variables are
measuring nearly the same thing and destabilising coefficient estimates.

```{r age-hist}
ggplot(hfc_records, aes(age, fill = DEATH_EVENT)) +
  geom_histogram(bins = 30) +
  labs(title = "Age distribution by survival outcome",
       subtitle = "Deaths (1) appear more concentrated at older ages",
       x = "Age (standardised)", y = "Count", fill = "Died") +
  theme_minimal()
```

The age histogram gives an early indication that older patients are over-represented in the
death group, consistent with age being a known risk factor in heart failure prognosis.
However, the overlap between groups is substantial — age alone will not be sufficient to
classify patients accurately.

---

## Train / Test Split

```{r Train Test}
train_set <- createDataPartition(hfc_records$DEATH_EVENT, p = 0.8, list = FALSE)
train <- hfc_records[train_set, ]
test  <- hfc_records[-train_set, ]
```

We partition the data into **80% training (~239 patients)** and **20% test (~60 patients)**.
`createDataPartition` stratifies by the outcome variable, which preserves the 67/33 class
ratio in both partitions. The test set is sealed from all modelling decisions — it is only
used once at the end to report final performance.

With only 299 patients total, this is a small dataset. Small samples mean performance
estimates from a single split can be noisy — a handful of patients in the test set landing
on one side or the other of the boundary can swing accuracy by several percentage points.
This is precisely why we use **10-fold cross-validation** rather than relying on the split
alone.

---

## Logistic Regression

Logistic regression models the *probability* that a patient dies, as a function of their
clinical measurements. It is the standard first-line tool for binary clinical outcomes because
its coefficients have a direct interpretation: each coefficient represents the change in the
log-odds of death for a one-unit increase in that predictor, holding all others constant.

```{r Logistic Regression}
log_model <- glm(DEATH_EVENT ~ ., data = train, family = binomial)
summary(log_model)
```

The full model includes all 12 predictors. Looking at the summary, predictors with small
p-values (conventionally below 0.05) are those showing the strongest independent associations
with mortality after controlling for all other variables. Notably, **ejection fraction**,
**serum creatinine**, and **time** (follow-up duration) typically emerge as significant — 
consistent with the clinical literature on heart failure prognosis.

The `time` variable deserves a note: it represents how long the patient was followed up,
which partly reflects survival itself. In a prospective clinical model this would require
careful handling, but for this predictive exercise we retain it.

```{r stepwise}
# Stepwise AIC reduces the model to a smaller set of predictors.
# AIC (Akaike Information Criterion) penalises model complexity — adding a predictor
# only "pays off" if it improves fit enough to justify the added parameter.
# The "both" direction means variables can be added or removed at each step.
step_model <- step(log_model, direction = "both")
summary(step_model)
```

The stepwise procedure has reduced the model to 5 key features. This is useful for
interpretability, but it is a greedy sequential process — it does not guarantee finding the
globally best subset, and the selected variables can change substantially with small changes
to the data. We compare this to LASSO below, which is a more principled approach to variable
selection.

---

## Cross-Validation Setup

```{r cv-setup}
train_control <- trainControl(method = "cv",
                              number = 10,
                              classProbs   = TRUE,
                              summaryFunction = twoClassSummary)

# Relabel factor levels to valid R names — required by caret's internal functions
levels(train$DEATH_EVENT) <- c("No", "Yes")
levels(test$DEATH_EVENT)  <- c("No", "Yes")
```

We use **10-fold cross-validation** for all models. The training data is divided into 10
equal chunks; each model is trained 10 times, each time using 9 chunks for training and
1 chunk for validation, rotating through all 10. The 10 validation scores are averaged
to give a stable performance estimate.

This is more reliable than a single train/test split, especially with a small dataset like
this one. It uses the training data more efficiently while still providing an honest
estimate of how the model will perform on data it has never seen.

`twoClassSummary` tells caret to report **AUC-ROC** as the primary metric, rather than
simple accuracy. AUC-ROC is preferred for imbalanced classification problems because it
simultaneously evaluates sensitivity and specificity across all possible classification
thresholds — it is robust to the class imbalance in this dataset in a way that accuracy
is not.

---

## LDA and QDA

```{r lda-qda}
lda_model <- lda(DEATH_EVENT ~ ., data = train)
lda_pred  <- predict(lda_model, test)$class

qda_model <- qda(DEATH_EVENT ~ ., data = train)
qda_pred  <- predict(qda_model, test)$class
```

**LDA (Linear Discriminant Analysis)** finds the linear combination of predictors that best
separates the two outcome classes. It assumes that both classes have the same covariance
structure — i.e. the spread and shape of the predictor distributions is the same for
survivors and deaths, just centred in different places.

**QDA (Quadratic Discriminant Analysis)** relaxes this assumption and estimates a separate
covariance matrix for each class, allowing the decision boundary between classes to be curved
rather than a straight line. The trade-off is that QDA requires estimating substantially more
parameters — with only 239 training patients and 12 predictors, this is a concern. A model
that needs to estimate more parameters than the data comfortably supports will tend to overfit.

---

## LASSO Variable Selection

LASSO (Least Absolute Shrinkage and Selection Operator) selects predictors by applying a
penalty to the size of coefficients. As the penalty increases, less important predictors are
shrunk to exactly zero and effectively removed from the model. The optimal penalty level is
chosen by cross-validation — so the variable selection itself is cross-validated, not just
the model performance.

This is methodologically preferable to stepwise AIC because it optimises shrinkage and
selection simultaneously rather than greedily stepping through one variable at a time.

```{r LASSO}
x_train <- model.matrix(DEATH_EVENT ~ ., data = train)[, -1]
y_train <- train$DEATH_EVENT

# CV LASSO — finds the penalty (lambda) that minimises cross-validated error
lasso_cv <- cv.glmnet(x_train, y_train, family = "binomial", alpha = 1)
plot(lasso_cv)
```

This plot shows the cross-validated error (binomial deviance, y-axis) across the range of
penalty values (log lambda, x-axis). The left dashed line marks `lambda.min` — the penalty
that minimises error. The right dashed line marks `lambda.1se` — a slightly larger penalty
that gives a simpler model within one standard error of the minimum error. We use `lambda.1se`
to favour parsimony.

```{r lasso-coefs}
lasso_coefs  <- coef(lasso_cv, s = "lambda.1se")
selected_vars <- rownames(lasso_coefs)[which(lasso_coefs != 0)]
selected_vars <- selected_vars[selected_vars != "(Intercept)"]
print(selected_vars)

lasso_formula <- as.formula(paste("DEATH_EVENT ~", paste(selected_vars, collapse = " + ")))
```

LASSO has reduced the model to **3 variables**. The fact that both the stepwise AIC model
(5 variables) and LASSO (3 variables) converge on a small subset of the original 12 predictors
is meaningful — it suggests the remaining predictors carry little additional signal once the
key variables are accounted for. Agreement between two different selection methods on the same
small set increases confidence that these are genuinely the most important clinical markers.

```{r lasso-path}
# The coefficient path shows how each predictor's coefficient changes as the
# penalty increases from right (no penalty, all variables included) to left
# (high penalty, most variables zeroed out). Variables that survive to the far
# left are the most robustly important.
lasso_path <- glmnet(x_train, y_train, family = "binomial", alpha = 1)

plot(lasso_path, xvar = "lambda", label = TRUE)
abline(v = log(lasso_cv$lambda.min), lty = 2, col = "blue")
abline(v = log(lasso_cv$lambda.1se), lty = 2, col = "red")
legend("topright", legend = c("lambda.min", "lambda.1se"),
       col = c("blue", "red"), lty = 2)
```

Each line in this plot is one predictor. Variables that remain non-zero all the way to the
right dashed line (`lambda.1se`) are those retained in our final LASSO model — they are the
predictors that survive even aggressive penalisation, indicating they carry the most
independent predictive signal.

---

## Cross-Validated Model Comparison

Now we train all four models under identical 10-fold CV conditions so performance comparisons
are fair — every model sees the same folds.

```{r cv-models}
cv_logistic <- train(
  DEATH_EVENT ~ ., data = train,
  method    = "glm",
  family    = "binomial",
  trControl = train_control,
  metric    = "ROC"
)

cv_logistic_lasso <- train(
  lasso_formula, data = train,
  method    = "glm",
  family    = "binomial",
  trControl = train_control,
  metric    = "ROC"
)

cv_lda <- train(
  DEATH_EVENT ~ ., data = train,
  method    = "lda",
  trControl = train_control,
  metric    = "ROC"
)

cv_qda <- train(
  DEATH_EVENT ~ ., data = train,
  method    = "qda",
  trControl = train_control,
  metric    = "ROC"
)
```

```{r cv-results}
results <- resamples(list(
  Logistic       = cv_logistic,
  Logistic_LASSO = cv_logistic_lasso,
  LDA            = cv_lda,
  QDA            = cv_qda
))

summary(results)
dotplot(results, metric = "ROC")
```

The dot plot shows each model's cross-validated AUC-ROC with 95% confidence intervals.
AUC-ROC ranges from 0.5 (no better than random guessing) to 1.0 (perfect discrimination).

Key observations from the CV results:

- **Logistic LASSO and LDA** achieve the highest point estimates, with wide but overlapping
  confidence intervals — given the small sample size, we cannot confidently declare one
  superior to the other.
- **QDA** shows the lowest and most variable performance. This is consistent with the concern
  raised earlier: estimating separate covariance matrices per class requires more data than
  this dataset provides, and the result is an unstable model.
- The confidence intervals are wide across all models, a direct consequence of the small
  sample. This is an important caveat — with only 60 test patients, a difference of one or
  two correct classifications can change accuracy by several percentage points.

---

## Final Test Set Evaluation

```{r final-eval}
x_test <- model.matrix(DEATH_EVENT ~ ., data = test)[, -1]

preds <- data.frame(
  Logistic       = predict(cv_logistic,       test),
  Logistic_LASSO = predict(cv_logistic_lasso, test),
  LDA            = predict(cv_lda,            test),
  QDA            = predict(cv_qda,            test)
)

lapply(preds, function(p) confusionMatrix(p, test$DEATH_EVENT, positive = "Yes"))
```

Reading the confusion matrices — for each model there are four possible outcomes for any
given patient:

- **True Positive (TP)**: predicted to die, actually died — correct, clinically important
- **True Negative (TN)**: predicted to survive, actually survived — correct
- **False Negative (FN)**: predicted to survive, actually died — the most costly error
- **False Positive (FP)**: predicted to die, actually survived — causes unnecessary concern

In a clinical mortality prediction context, **sensitivity** (the proportion of actual deaths
correctly identified) is the priority metric — missing a patient who will die is a more
serious error than a false alarm. We focus on this when comparing models.

The **logistic regression (full) and LDA produce identical confusion matrices** — this is
not a coincidence. LDA and logistic regression are asymptotically equivalent when the
predictor distributions are approximately normal and the classes share the same covariance
structure. Their convergence here suggests those assumptions hold reasonably well, and that
a linear decision boundary is appropriate for this data.

**QDA's sensitivity is notably lower** than the other models — it misclassifies a
disproportionate number of actual deaths as survivors. Despite its higher specificity
(it is good at identifying survivors), this trade-off is unfavourable for a clinical risk
screening tool.

```{r roc-curve}
prob_log <- predict(cv_logistic, test, type = "prob")[, 2]
roc_log  <- roc(test$DEATH_EVENT, prob_log)

plot(roc_log, main = "ROC Curve — Logistic Regression (full model)",
     col = "steelblue", lwd = 2)
auc(roc_log)
```

The ROC curve plots sensitivity against (1 - specificity) across all possible classification
thresholds. The default threshold of 0.5 — predicting "death" if the model assigns >50%
probability — is one point on this curve. Depending on the clinical context, a different
threshold might be preferred: if false negatives (missed deaths) are especially costly,
lowering the threshold increases sensitivity at the expense of more false alarms.

The **AUC** (Area Under the Curve) summarises overall discriminative performance as a single
number. An AUC of 0.85 means that for a randomly selected pair of one patient who died and
one who survived, the model correctly assigns the higher probability to the patient who died
85% of the time.

---

## Summary of Findings

### Model performance

| Model             | Accuracy | Sensitivity | Specificity | Balanced Acc. |
|-------------------|----------|-------------|-------------|----------------|
| Logistic (full)   | 84.7%    | 73.7%       | 90.0%       | 81.8%          |
| Logistic (LASSO)  | 79.7%    | 63.2%       | 87.5%       | 75.3%          |
| LDA               | 84.7%    | 73.7%       | 90.0%       | 81.8%          |
| QDA               | 78.0%    | 47.4%       | 92.5%       | 69.9%          |

The full logistic model and LDA are the best-performing methods on this dataset, achieving
a balanced accuracy of 81.8% — meaning they are correct roughly 4 times out of 5 when
accounting equally for both outcome classes.

**LASSO's underperformance relative to the full logistic model** is likely a consequence of
the conservative `lambda.1se` criterion over-shrinking in a small-sample setting. With only
239 training patients, removing variables that carry even modest signal can noticeably hurt
performance. Using `lambda.min` instead would retain more variables and likely recover some
of this gap.

**QDA's poor sensitivity (47%)** means it misses more than half of actual deaths — a
clinically unacceptable failure rate for a risk stratification tool. This reinforces the
earlier concern about estimating class-specific covariance structures with insufficient data.

### Clinical interpretation

The variables consistently selected by both LASSO and stepwise AIC — most notably
**serum creatinine**, **ejection fraction**, and **follow-up time** — align with established
clinical knowledge. Serum creatinine is a marker of kidney function, which deteriorates in
advanced heart failure. Ejection fraction directly measures the heart's pumping efficiency.
Their emergence as the dominant predictors, across multiple independent selection methods,
is reassuring that the models are capturing genuine clinical signal rather than noise.

### Limitations

- The dataset contains only 299 patients from a single centre (Faisalabad Institute of
  Cardiology, Pakistan, 2015). Performance on populations with different demographics or
  care standards may differ.
- The small sample produces wide confidence intervals around all performance estimates —
  the exact figures should be treated as indicative rather than precise.
- All models use the default 0.5 classification threshold. In practice, this threshold
  should be adjusted based on the relative costs of false negatives versus false positives
  in the specific clinical setting — a threshold of 0.3 would increase sensitivity at the
  cost of more false alarms, which may be the right trade-off for a screening tool.
- This is an observational dataset. The associations identified do not establish causality —
  they indicate which measurements are *predictive* of mortality, not necessarily which
  factors *cause* it.
